Balsa: A Compact Line Notation Based on SMILES

### 2022-11-01 manuscript.tex
Fixes typos.

Balsa's main goal is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. This allows general-purpose devices and networks to convey complex domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure. Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile (\enquote{molfile})~\cite{ctfileFormats}. Figure~\ref{fig:representative-examples} presents some simple chemical structures together with their Balsa serializations.

@@

Balsa's rules can be divided into two broad categories: \textit{syntax} and \textit{semantics}. Syntax defines the set of grammatically valid strings for a language, whereas semantics determines their meaning. Balsa semantic are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.

@@

\begin{itemize}
\item \textit{Constitution}, the atoms present in a molecule and their bonding relationships. Balsa's extends the valence bond model with a limited form of delocalized bonding.
\item \textit{Conformation}, a kind of stereoisomerism resulting from restricted rotation about a bond. Balsa supports restricted rotation about double bonds capable of producing alkene isomerism.
\item \textit{Configuration}, a kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space. Balsa supports configurations for atoms having tetrahedral symmetry attached to four neighboring atoms.
\end{itemize}

@@

The main advantage of trees over graphs is information density. Graphs must accommodate cycles, a requirement fulfilled by assigning a unique identifier to each node. Each edge then references two of these identifiers. Graph representations therefore incur the storage overhead of identifiers associated with both nodes and edges. In contrast, a tree can be directly encoded as a set of nested, alternating paths of nodes and edges. This simplifies the representation and eliminates the overhead of assigning and using node identifiers.

@@

It is sometimes convenient to interconvert virtual and atomic hydrogens. The transformation of an atomic hydrogen into a virtual hydrogen is called \textit{reification}. The reverse process is called \textit{virtualization}.

The \ttt{charge} (\enquote{charge}) and \ttt{selected} (\enquote{selected}) attributes determine valence electron count (see: \nameref{electron-counting}). Charge refers formal charge, or the difference between the valence electron count of the element and the sum of bonding and nonbonding electrons of the bound atom. By default, an atom is assigned a charge of zero. Minimum and maximum values for charge are -9 and +9, respectively. The \ttt{selected} attribute adjusts an atom's selection status. Selecting an atom allows it to participate in extended bonding (see: \nameref{delocalization-subgraph}).

@@

Electron counting in Balsa is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. This model assumes that a bond spans exactly two atoms and is associated with a nonzero, positive, and even electron count. These electrons are drawn in equal quantities from each atom's set of valence electrons. A single bond draws one electron from each atom (two total), a double bond two (four total), and a triple bond three (six total). The electron count of an atom whose \ttt{element} attribute equals \ttt{None} is always zero. Bonding does not change it.

@@

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME, Figure~\ref{fig:dime}). DIME results from limitations to the valence bond model causing two representations to encode what would be considered the same molecule. When the two forms differ only in double bond placement, the terms \enquote{resonance} or \enquote{aromaticity} are sometimes used. These terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

Special handling is required for \textit{undefined stereocenters}. A stereocenter is undefined if it lacks a parity. Undefined stereocenters indicate that no information about a central atom's configuration is known. This could indicate the presence of either a single configuration or a configurational mixture. This interpretation is consistent with the one used by molfile format~\cite{ctfileFormats}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{gratuitous-selection.pdf}
    \caption{Gratuitous selection. Heteroatoms are selected needlessly because subvalence is zero in every case. Atoms selected gratuitously, must be pruned before processing the DS.}
    \label{fig:gratuitous-selection}
\end{figure}

@@

Writers should consider the non-negligible costs of atom selection. Algorithms for selection are likely to involve the perception of cycles, and so could exhibit superlinear time complexity. Often, a reader must perform a global deselection to arrive at a localized valence bond representation, which at the very least requires pruning and a maximal matching procedure. In other words, atom selection imposes two sets of costs: one on the writer and the other on every subsequent reader forever into the future.

@@

These metrics compare favorably with alternatives. Consider the molfile (V2000) format. 32 characters per atom are required in the best case, and 51 characters in the worst. All bonds, regardless of type, require between nine and 12 characters. Atomic charges and isotopes typically require additional characters.

@@

The most noteworthy feature of Balsa is its compatibility with SMILES. Contemporary SMILES implementations will be able to read and write Balsa strings --- at least to a point. Incompatibilities will arise from three main sources: (1) those features of SMILES that Balsa has deliberately omitted; (2) those aspects of the SMILES language that are incompletely-specified, ambiguous, or self-contradictory; and (3)those features of SMILES that were deliberately disregarded by implementations.

@@

This paper describes Balsa, a compact molecular serialization format designed to be forward compatible with SMILES. Constitutionally, Balsa can encode any molecule conforming to the valence bond model. In the event of undesired symmetry artifacts due to delocalization, Balsa offers a mitigation based in graph theory. Conformational isomerism of alkenes is supported by partial parity bonds. The configurations of tetracoordinate, stereogenic atoms are encoded through the use of a parity enumeration and conventions around its use.

>>>>

Balsa's main goal is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. This allows general-purpose devices and networks to convey complex domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure. Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile (e.g., \enquote{Molfile})~\cite{ctfileFormats}. Figure~\ref{fig:representative-examples} presents some simple chemical structures together with their Balsa serializations.

@@

Balsa's rules can be divided into two broad categories: \textit{syntax} and \textit{semantics}. Syntax defines the set of grammatically valid strings for a language, whereas semantics determines their meaning. Balsa semantics are complex due to the brevity of the syntax and the inherently complex nature of molecular representation.

@@

\begin{itemize}
\item \textit{Constitution}, the atoms present in a molecule and their bonding relationships. Balsa's extends the electron-pair bonding model with a limited form of delocalized bonding.
\item \textit{Conformation}, a kind of stereoisomerism resulting from restricted rotation about a bond. Balsa supports restricted rotation about double bonds capable of producing alkene isomerism.
\item \textit{Configuration}, a kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space. Balsa supports configurations for atoms having tetrahedral symmetry attached to four neighboring atoms.
\end{itemize}

@@

The main advantage of trees over graphs is information density. Graphs must accommodate cycles, a requirement fulfilled by assigning a unique identifier to each node. Each edge then references two of these identifiers. Graph representations therefore incur the storage overhead of identifiers associated with both nodes and edges. In contrast, a tree can be directly encoded as a set of nested, alternating paths of nodes and edges. This simplifies representation and eliminates the overhead of assigning and using node identifiers.

@@

It is sometimes convenient to interconvert virtual and atomic hydrogens. The transformation of an atomic hydrogen into a virtual hydrogen is called \textit{virtualization}. The reverse process is called \textit{reification}.

The \ttt{charge} (\enquote{charge}) and \ttt{selected} (\enquote{selected}) attributes determine valence electron count (see: \nameref{electron-counting}). Charge refers to formal charge, or the difference between the valence electron count of the element and the sum of bonding and nonbonding electrons of the bound atom. By default, an atom is assigned a charge of zero. Minimum and maximum values for charge are -9 and +9, respectively. The \ttt{selected} attribute adjusts an atom's selection status. Selecting an atom allows it to participate in extended bonding (see: \nameref{delocalization-subgraph}).

@@

Electron counting in Balsa is based on the well-known electron-pair bonding model developed by Lewis and others (\enquote{Lewis model})~\cite{lewis:1916}. This model assumes that a bond spans exactly two atoms and is associated with a nonzero, positive, and even electron count. These electrons are drawn in equal quantities from each atom's set of valence electrons. A single bond draws one electron from each atom (two total), a double bond two (four total), and a triple bond three (six total). The electron count of an atom whose \ttt{element} attribute equals \ttt{None} is always zero. Bonding does not change it.

@@

A molecular representation based solely on the Lewis model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME, Figure~\ref{fig:dime}). DIME results from limitations of the Lewis model causing two representations to encode what would be considered the same molecule. When the two forms differ only in double bond placement, the terms \enquote{resonance} or \enquote{aromaticity} are sometimes used. These terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

Special handling is required for \textit{undefined stereocenters}. A stereocenter is undefined if it lacks a parity. Undefined stereocenters indicate that no information about a central atom's configuration is known. Either a single configuration or a configurational mixture is present. This interpretation is consistent with the one used by Molfile format~\cite{ctfileFormats}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{gratuitous-selection.pdf}
    \caption{Gratuitous selection. Heteroatoms are selected needlessly because subvalence is zero in every case. Atoms selected gratuitously must be pruned before processing the DS.}
    \label{fig:gratuitous-selection}
\end{figure}

@@

Writers should consider the non-negligible costs of atom selection. Algorithms for selection are likely to involve the perception of cycles, and so could exhibit superlinear time complexity. Often, a reader must perform a global deselection to arrive at a localized Lewis representation, which at the very least requires pruning and a maximal matching procedure. In other words, atom selection imposes two sets of costs: one on the writer and the other on every subsequent reader forever into the future.

@@

These metrics compare favorably with alternatives. Consider the Molfile (V2000) format. 32 characters per atom are required in the best case, and 51 characters in the worst. All bonds, regardless of type, require between nine and 12 characters. Atomic charges and isotopes typically require additional characters.

@@

The most noteworthy feature of Balsa is its compatibility with SMILES. Contemporary SMILES implementations will be able to read and write Balsa strings --- at least to a point. Incompatibilities will arise from three main sources: (1) those features of SMILES that Balsa has deliberately omitted; (2) those aspects of the SMILES language that are incompletely-specified, ambiguous, or self-contradictory; and (3) those features of SMILES that were deliberately disregarded by implementations.

@@

This paper describes Balsa, a compact molecular serialization format designed to be forward compatible with SMILES. Constitutionally, Balsa can encode any molecule conforming to the electron-pair bonding model. In the event of undesired symmetry artifacts due to delocalization, Balsa offers a mitigation based in graph theory. Conformational isomerism of alkenes is supported by partial parity bonds. The configurations of tetracoordinate, stereogenic atoms are encoded through the use of a parity enumeration and conventions around its use.
